A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-03 paint-contour-tree.tex
added changes to intro

\begin{abstract}
Consider a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a simplicial mesh in $\RR^d$. 
A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
As $v$ is changed, these contours change topology, merge into each other, or split.
\emph{Contour trees} are concise representations of $f$ that track this contour
behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
The edges represent changes in the topology of contours.
It is a fundamental data structure in data analysis and visualization,
and there is significant previous work (both theoretical and practical) on algorithms for constructing contour trees.

Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
A classic result of Carr, Snoeyink, and Axen (2000) gives an algorithm that takes $O(n\log n + N\alpha(N))$ time
(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
was given by Chiang et al. All these algorithms involve a global sort of the critical points,
which is a major bottleneck for large datasets. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
always bounded by $O(t\log t + N)$ and is $O(N+t\alpha(t))$ for balanced trees. 
Moreover, we prove a general lower bound of
$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
that the tree structure must play a role in the time complexity.

@@

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold 
(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.

@@

Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
$\MM$ in $\RR^{d+1}$ with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. These are the relevant events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical points},
where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
representing the important topological events. An edge connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.)

@@

But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.

@@

\begin{theorem} \label{thm:main-alg} Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
This is clearly an improvement when the tree is short and fat.
In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case, and fix a mesh $\MM$.
As referenced in \Thm{main-alg}, for any height function $f$ our algorithm defines a 
specific path decomposition $P(T)$\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}. 
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$ 
defined over the fixed mesh $\MM$\footnote{If one is concerned
that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
a refined lower bound matching \Thm{main-alg}.

@@

\begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$ 
(defined over a fixed mesh $\MM$).  Any height function decision tree that correctly computes
the contour tree on all of $\bF_C$ has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
\end{theorem}

@@

In other words, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
of inputs of fixed cost.

>>>>

\begin{abstract}
Consider a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
As $v$ is changed, these contours change topology, merge into each other, or split.
\emph{Contour trees} are concise representations of $f$ that track this contour
behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
The edges represent changes in the topology of contours.
It is a fundamental data structure in data analysis and visualization,
and there is significant previous work (both theoretical and practical) on algorithms for constructing contour trees.

Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
A classic result of Carr, Snoeyink, and Axen (2000) gives an algorithm that takes $O(n\log n + N\alpha(N))$ time
(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
was given by Chiang et al. All these algorithms involve a global sort of the critical points,
a significant computational bottleneck. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
always bounded by $O(t\log t + N)$ and is $O(t\alpha(t)+N)$ for balanced trees. 
Moreover, we prove a strong optimality property of our algorithm, showing that $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. Consider the family of instances where our algorithm
requires $C$ comparison. Any algorithm computing contour trees on this family requires
$\Omega(C)$ comparisons in the worst case.

%
%general lower bound of
%$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
%that the tree structure must play a role in the time complexity.

@@

Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
and in doing so yields instance optimal results.
Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
exists some partition $P'$ such that the running time of our algorithm is 
$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold.
In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.

@@

Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
$\MM$ in $\RR^{d+1}$ with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours may acquire or lose handles. These are the relevant events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
An edge connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.)

@@

But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for some instances? Is there a more refined version of the complexity
of computing contour trees, and can we characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.

@@

\begin{theorem} \label{thm:main-alg} Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is a deterministic algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
This is clearly an improvement when the tree is short and fat.
But it can be even better. A calculation yields that for any decomposition of a rooted balanced binary tree, 
our bound is $O(t\alpha(t) + n)$. 
%
%
%In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
%However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
specific path decomposition $P(T)$\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}. 
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for ant $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
a refined lower bound matching \Thm{main-alg}.

@@

\begin{theorem} \label{thm:main-lb} 
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider the set of functions $\bF_C$ that all same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

@@

Roughly speaking, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
of inputs of fixed cost.
